---
# UNL thesis fields
title: "HUMAN PERCEPTION OF EXPONENTIALLY INCREASING DATA DISPLAYED ON A LOG SCALE"
author: "Emily Anna Robinson"
month: "August"
year: "2022"
location: "Lincoln, Nebraska"
major: "Statistics"
adviser: "Susan VanderPlas and Reka Howard"
adviserAbstract: 
abstract: |
  Log scales are often used to display data over several orders of magnitude within one graph. During the COVID-19 pandemic, we have seen both the benefits and the pitfalls of using log scales to display case counts. Three graphical experimental tasks were conducted to evaluate the impact our choice of scale has on human perception of exponentially increasing trends. The first experiment evaluates whether our ability to perceptually notice differences in exponentially increasing trends is impacted by the choice of scale. We conducted a visual inference experiment in which participants were shown a series of lineup plots (consisting of 19 null panels and 1 target panel) and asked to identify the panel that was most different from the others. Our results indicated that when there was a large difference in curvature between the target plot and null plots, the choice of scale had no impact and participants accurately differentiated between the two curves on both the linear and log scale. However, displaying exponentially increasing data on a log scale improved the accuracy of differentiating between models with slight curvature differences. An exception occurred when identifying a plot with curvature embedded in surrounding plots closely relating to a linear trend, indicating that it is easy to identify a curve in a group of lines but much harder to identify a line in a group of curves. The use of visual inference to identify these guidelines suggests that there are *perceptual* advantages to log scales when differences are subtle. Our other experimental tasks focus on determining whether there are cognitive disadvantages to log scales: do log scales make it harder to make use of graphical information? We conducted a graphical task similar to the New York Times "You Draw It" page to test an individual's ability to use and make predictions for exponentially increasing data. We asked participants to draw a line using their computer mouse through the increasing exponential trend shown on both scales. In addition to differentiation and prediction of exponentially increasing data, we conduct an experimental task to test an individuals' ability to translate a graph of exponentially increasing data into real value quantities and extend their estimations by making comparisons. The results of our experimental tasks allow us to provide guidelines for readers to actively choose which of many possible graphics to draw, according to some set of design choices, to ensure that our charts are effective. *(399 words; 350 word limit)*
acknowledgments: |
  Thank you to all my people!
dedication: |
  Dedicated to...
# End of UNL thesis fields
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    pandoc_args: --top-level-division=chapter
    keep_tex: yes
    latex_engine: xelatex
    template: template.tex
  huskydown::thesis_gitbook: 
    style: style.css
#  huskydown::thesis_word: default
#  huskydown::thesis_epub: default
bibliography: bib/thesis.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: bib/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#header-includes:
#- \usepackage{tikz}
---

```{r setup, include = F}
options(width = 60)
knitr::opts_chunk$set(
  echo = F, eval = T, messages = F, warnings = F,
  fig.width = 6, fig.height = 4,  fig.align = 'center',
  out.width = "\\linewidth", dpi = 300, 
  tidy = T, tidy.opts=list(width.cutoff=45)
)
```

```{r include_packages, include = FALSE}
# This chunk ensures that the huskydown package is
# installed and loaded. This huskydown package includes
# the template files for the thesis.
if(!require(devtools))
  install.packages("devtools", 
                   repos = "http://cran.rstudio.com")
if(!require(huskydown))
  devtools::install_github(
    "benmarwick/huskydown"
  )
library(huskydown)
library(readr)
library(tidyverse)
library(scales)
library(knitr)
library(gridExtra)
```

# Literature Review

Editing text colors: \ear{Emily's editing color.} Emily may also use mostly black text as well. \svp{Susan's editing color.} \rh{Reka's editing color.}

+ Introduction

With the advancement of technology, access to data is more readily available more than ever before.
With this abundance of data comes the responsibility to make sense out of the numbers in order to convey the message clearly. 
Graphics are a useful tool for displaying and communicating information [@vanderplas2020testing].
Researchers include graphics to communicate their results in scientific publications and news sources rely on graphics to convey news stories to the public [\ear{CITE ME}].

+ Motivation (COVID-19)
    + @lisa_charlotte_2020
    + @romano_scale_2020
    + @rost_2020
    + @bavel_using_2020
    + @wade_fagen_ulmschneider_2020
    + @financial_times_2020
    
At the beginning of the SARS-NCOV-2 pandemic (COVID-19), we saw an influx of dashboards being developed to display case counts, transmission rates, and outbreak regions [@lisa_charlotte_2020]; mass media routinely showed charts to share information with the public about the progression of the pandemic [@romano_scale_2020].
People began seeking out graphical displays of COVID-19 data as a direct result of these pieces of work [@rost_2020]; providing increased and ongoing exposure to these graphics over time.
Many of these graphics helped guide decision makers to implement policies such as shut-downs or mandated mask wearing, as well as facilitated communication with the public to increase compliance [@bavel_using_2020].

+ General Sensation and Perception
+ General Graphics Research
    + @vanderplas2020testing
    + @unwin_why_2020
    
With the increasing importance graphics play in our everyday lives, we must actively choose which of many possible graphics to draw, according to some set of design choices, to ensure that our charts are effective, as suggested in [@unwin_why_2020].

+ Logrithms
    + @menge_logarithmic_2018
    + @xkcd
    + @heckler_student_2013
    + @varshney_why_2013
    + @siegler_numerical_2017
    + @dehaeneLogLinearDistinct2008
+ Visual Inference through Lineups
    + @best_perception_2007
    + @vanderplas_clusters_2017
    + @hofmann_graphical_2012
    + @loyVariationsQQPlots2016
    + @wickham2010graphical
    + @majumder_validation_2013
    + @vanderplas_statistical_nodate
+ Prediction with You Draw It
    + @wagenaar_misperception_1975
    + @mackinnon_feedback_1991
    + @jones_polynomial_1977
    + @jones_generalized_1979
    + @wagenaar_extrapolation_1978
    + @timmers_inverse_1977
    + @mosteller_eye_1981
    + @finney1951subjective
+ Numerical Translation and Estimation

*Text below copied from JSM 2021 Student Paper*


When faced with data which spans several orders of magnitude, we must decide whether to show the data on its original scale (compressing the smaller magnitudes into relatively little area) or to transform the scale and alter the contextual appearance of the data.
One common solution is to use a log scale transformation to display data over several orders of magnitude within one graph.
Logarithms make multiplicative relationships additive, showing elasticities and other proportional changes, and also linearize power laws [@menge_logarithmic_2018]. 
When presenting log scaled data, it is possible to use either untransformed scale labels (for example, values of 1, 10 and 100 are equally spaced along the axis) or log transformed scale labels (for example, 0, 1, and 2, showing the corresponding powers of 10).
We have recently experienced the benefits and pitfalls of using log scales as COVID-19 dashboards displayed case count data on both the log and linear scale [@wade_fagen_ulmschneider_2020, @financial_times_2020]. 
In spring 2020, during the early stages of the COVID-19 pandemic, there were large magnitude discrepancies in case counts at a given time point between different geographic regions (e.g. states and provinces as well as countries and continents). 
During this time, we saw the usefulness of log scale transformations showing case count curves for areas with few cases and areas with many cases within one chart.
As the pandemic evolved, and the case counts were no longer spreading exponentially, graphs with linear scales seemed more effective at spotting early increases in case counts that signaled more localized outbreaks.
This is only one recent example of a situation in which both log and linear scales are useful for showing different aspects of the same data; there are long histories of using log scales to display results in ecology, psychophysics, engineering, and physics 
[@xkcd, @menge_logarithmic_2018, @heckler_student_2013].
In fact, research shows that we do not perceive exponential growth accurately (whether information is presented in graphical or tabular form): we systematically under-estimate exponential growth.

Research suggests our perception and mapping of numbers to a number line is logarithmic at first, but transitions to a linear scale later in development, with formal mathematics education [@varshney_why_2013, @siegler_numerical_2017, @dehaeneLogLinearDistinct2008].
This transition to linear scales occurs first in small numbers (e.g. 1-10) and then gradually expands to higher orders of magnitude; thus, the logarithmic intuition about numbers in children is often more noticeable on scales in the thousands to hundreds of thousands.
If we perceive logarithmically by default, it is a natural (and presumably low-effort) way to display information and should be easy to read and understand/use.
In fact, early studies explored the estimation and prediction of exponential growth, finding that growth is underestimated when presented both numerically and graphically but that numerical estimation is more accurate than graphical estimation for exponential curves [@wagenaar_misperception_1975].
One way to improve estimation of increasing exponential trends is to provide immediate feedback to participants about the accuracy of their current predictions [@mackinnon_feedback_1991].
While prior contextual knowledge or experience with exponential growth does not improve estimation, instruction on exponential growth reduces the underestimation: participants adjust their initial starting value but not their perception of growth rate [@wagenaar_misperception_1975, @jones_polynomial_1977].
Estimation was shown to improve when subjects were presented with decreasing exponential functions [@timmers_inverse_1977].
@jones_polynomial_1977,@jones_generalized_1979 and @wagenaar_extrapolation_1978 propose competing polynomial models for the perception and extrapolation of exponential series.
It seems that estimation is a two-stage process: first, we identify the type of curve and direction and then, we use that information for prediction [@best_perception_2007].

Our inability to accurately predict exponential growth might also be addressed by log transforming the data, however, this transformation introduces new complexities; most readers are not mathematically sophisticated enough to intuitively understand logarithmic math and translate that back into real-world effects.
In @menge_logarithmic_2018, ecologists were surveyed to determine how often ecologists encounter log scaled data and how well ecologists understand log scaled data when they see it in the literature. 
Participants were presented two relationships displayed on linear-linear scales, log-log scales with untransformed values, or log–log scales with log transformed values. 
@menge_logarithmic_2018 propose three types of misconceptions participants encountered when presented data on log-log scales: 
'hand-hold fallacy', 'Zeno's zero fallacy', and 'watch out for curves fallacies'. These misconceptions are a result of linear extrapolation assuming that a line in log-log space represents a line instead of the power law in linear-linear space. The study found that in each of these scenarios, participants were confident in their incorrect responses, indicating incorrect knowledge rather than a lack of knowledge.

The 'hand-hold falacy' stems from the misconception that steeper slopes in log-log relationships are steeper slopes in linear-linear space.
In fact, it is not only the slope that matters, but also the intercept and the location on the horizontal axis since a line in log-log space represents a power law in linear-linear space (i.e. linear extraploation).
Emerging from 'Zeno's zero fallacy' is the misconception that positively sloped lines in log-log space can imply a non-zero value of y when x is zero. This is never true as postively sloped lines in log-log space actually imply that y = 0 when x = 0.
This misconception again is a result of linear extrapolation assuming that a line in log-log space represents a line instead of the power law in linear-linear space.
The last misconception, 'watch out for curves fallacies' encompases three faults: (1) lines in log-log space are lines in linear-linear space, (2) lines in log-log space curve upward in linear-linear space, and (3) curves in log-log space have the same curvature in linear-linear space.
Linear extrapolation is again responsible for the first and third faults while the second fault is a result of error in thinking that log-log lines represent power laws (which are exponential relationships), and all exponential relationships curve upward; this is only true when the log-log slope is greater than one. @menge_logarithmic_2018 found that in each of these scenarios, participants were confident in their incorrect responses, indicating incorrect knowledge rather than a lack of knowledge.

In order to provide a set of principles to guide design choices, we must evaluate these design choices through the use of graphical tests. 
These tests may take many forms: identifying differences in graphs, reading information off of a chart accurately, using data to make correct real-world decisions, or predicting the next few observations. 
All of these types of tests require different levels of use and manipulation of the information presented in the chart.
